\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2020-03-25 k-planar.tex
Added table of contents
Added \listfiles
Small changes + V_i => L_i

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}

@@

\usepackage[mathlines]{lineno}
% \setlength{\linenumbersep}{2em}
\linenumbers
% \rightlinenumbers
% \linenumbers
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
    {\linenomath\csname old#1\endcsname}%
    {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

\allowdisplaybreaks
%\sloppy
%%%%%%%%%%
\makeatletter
\def\NAT@spacechar{~}
\makeatother
%%%%%%%%%%
\crefname{lem}{Lemma}{Lemmas}
\crefname{thm}{Theorem}{Theorems}
\crefname{cor}{Corollary}{Corollaries}
\crefname{conj}{Conjecture}{Conjectures}
\crefname{obs}{Observation}{Observations}
\crefname{prop}{Proposition}{Propositions}
\crefname{clm}{Claim}{Claims}
\crefname{openproblem}{Open Problem}{Open Problems}
\crefformat{equation}{(#2#1#3)}
\Crefformat{equation}{Equation #2(#1)#3}

@@

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and implicit graph encoding. We extend this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
\end{titlepage}

% \tableofcontents
% \newpage

%\setcounter{\page}{1}
\section{Introduction}
\label{Introduction}

@@

A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in V_i$ and $w\in V_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.

@@

Apex-minor-free graphs, addressed by \cref{ApexMinorFree}, are the largest minor-closed class that admit bounded layered partitions~\citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

@@

\begin{lem}\label{generalized-tripod}
  Let $G$ be a graph; let $\mathcal{L}:=\langle V_1,\ldots,V_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$ where $H$ has treewidth at most $t$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \lemref{nice-decomposition}.  For each $x\in V(T)$, let $V_x := \bigcup_{y\in V(T_x)} Y_y$, $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x:=V_x\cup F_x$.  Then, 
  \begin{compactenum}[(Y1)]
    % \item\ylabel{thickness} $\mathcal{Y}=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width at most $\ell$ with respect to $\mathcal{L}$.
    \item\ylabel{separator} For each $x\in V(T)$, there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{lem}

@@

Next we show that $\mathcal{S}$ has small layered width with respect to the layering $\mathcal{L}=\langle V_1,\ldots,V_h\rangle$:

\begin{clm}\clmlabel{general-width}
  For each $i\in\{1,\ldots,h\}$ and each $x\in V(J)$, $|S_x\cap V_i|\le d\ell(k^2+3)$.  
\end{clm}

\begin{proof}
  We count the number of vertices in $S_x\cap V_i$ by upper-bounding the number of vertices contributed to $S_x\cap V_i$ by each vertex $w\in Y_x$. Refer to \cref{contribute}.  If $w\in Y_x\cap V_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap V_i$.

  \begin{figure}[htbp]
    \begin{center}
      \includegraphics{figs/contribute}
    \end{center}
    \caption{A path $P$ containing an internal vertex $w\in Y_x\cap V_{i-j}$.}
    \label{contribute}
  \end{figure}

  Otherwise, consider some path $P\in\mathcal{P}$ that contains $w$ as an internal vertex.  If $w\in V_{i}$, then $P$ contributes at most $k+1$ vertices to $S_x\cap V_i$.  If $w\in V_{i-1}\cup V_{i+1}$, then $P$ contributes at most $k$ vertices to $S_x\cap V_i$. If $w\in V_{i-j}\cup V_{i+j}$ for $j\ge 2$, then $P$ contributes at most $k-j$ vertices to $S_x\cap V_i$.

  For any $j$, the number of vertices $w\in V_{i+j}\cap Y_x$ is at most $\ell$. Each such vertex $w$ is an internal vertex of at most $d$ paths in $\mathcal{P}$. Therefore, 
  \[  |S_x\cap V_i|\le d\ell\cdot\left(k+1 + 2k + \sum_{j=2}^k 2(k-j)\right) %= d(2k+1) + \sum_{i=1}^{k-2} i
      = d\ell(k^2 +3) \enspace . \qedhere
  \]
\end{proof}

@@

  At this point, the proof of \cref{modern-major-general} is almost immediate from \clmref{general-width} and \clmref{general-bag-size}, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain edges $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+3)$ to $d\ell(k^3+3k)$.  Therefore $G$ has an $H$-partition of layered width at most $d\ell(k^3+3k)$ in which $H$ has treewidth at most $\binom{k+t}{t}$, completing the proof of \cref{modern-major-general}.
\end{proof}

@@

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, each path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$ contributes: (i)~at most two vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}\}$; (ii)~at most one vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2}+1,\ldots,k+1\}$; or (iii)~no vertices to $S_x\cap V_j$ for $j > k+1$.
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields 
\begin{align*}
    |S_x\cap Y_i| \le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left\lceil\frac{k+1}{2}\right\rceil
       \right)
       & =
       d\ell\left(
         2 + 2(k+1) + 2\left\lfloor\frac{k+1}{2}\right\rfloor
       \right) \\
       & \le
       d\ell(3k+5) 
      = 18k+30 \enspace .
\end{align*}    
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $(18k+30)(k+1)=18k^2+48k+30$.  This establishes \cref{k-planar}.
% Note to self-and-others. In our original SODA submission we had 18k^2+30k here because we forgot that the layer compression step has compress k+1 consecutive layers, not just k consecutive layers.

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analagous result for plane graphs \cite{dujmovic.joret.ea:planar}.
\begin{lem}
\label{induction} The setup:
  \begin{compactenum}
    \item Let $G$ and $G'$ be defined as above.
    \item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
    \item For each integer $j\ge 0$, let $V_j=\{v\in V(G):\dist_T(r,v)=j\}$. 
    \item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
    \begin{compactenum} 
      \item No edge of $F$ is crossed by any edge of $G$; and
      \item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $F[P_i]$ is a path; and
        \item $|V(P_i)\cap V_j| \le 15$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
  \end{compactenum}
  Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
  \begin{compactenum}
    \item $H$ is planar;
    \item for all integers $j\ge 0$, and all $x\in V(H)$, $|S_x\cap V_j|\le 15$; 
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$; and
    \item $H$ has a tree decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

@@

  We claim that, for each integer $i\ge 0$, $|V(\overline{Y}^+)\cap V_i|\le 15$.  First observe that, since $Q_1,Q_2,Q_3$ are each vertical paths in $T$,  $\overline{Y}$ contains at most three vertices of $V_i$, each incident on at most two edges of $\overline{Y}$.  Since $\dist_{G'}(v,w)\le 2$ for each $vw\in E(G)$, any vertex $x\in V(\overline{Y}^+)\setminus V(\overline{Y})\cap V_i$, is incident to an edge $xy\in E(G)$ that crosses one of the at most six edges in $\overline{Y}$ having an endpoint in $V_i$.  These at most six edges have at most 12 endpoints.  Therefore $|V(\overline{Y}^+)\setminus V(\overline{Y})\cap V_i|\le 6\times 2=12$, so $|V(\overline{Y}^+)\cap V_i|\le 12+3=15$.

@@

  Use the layering $\mathcal{L}=\langle V_0',V_1'\ldots\rangle$ where $V_i'=V_{2i}\cup V_{2i+1}$ for each integer $i\ge 0$. That this is a layering of $G$ follows from the fact that $\dist_{G'}(v,w)\le 2$ for every edge $vw\in E(G)$.  Since $|V_j\cap S_x|\le 15$ for every integer $j\ge 0$, $|V_i'\cap S_x|\le 30$ for every integer $i\ge 0$ and every $x\in V(H)$.
\end{proof}

% Exactly the same argument used to establish \cref{cor-k-planar} as a consequence of \cref{k-planar} establishes the following result as a consequence of \cref{1-planar}:
% 
% \begin{cor}
% \label{cor-1-planar}
% Let $G$ be a $1$-plane graph, let $r_0$ be an arbitrary vertex of $G$, let $G^+$ be a graph obtained by adding a dummy vertex at each crossing in $G$ and then arbitrarily adding non-crossing edges to obtain a triangulation.  Then $G$ has a layered $H$-partition $(\mathcal{L}=\langle V_0',V_1',\ldots\rangle,\mathcal{P})$ satisfying the conditions of \cref{1-planar} and such that $V_i' = \{v\in V(G): \floor{\dist_{G^+}(r_0,v)/2}=i\}$ for each integer $i\ge 0$.
% \end{cor}

@@

%Let $G$ be a $k$-plane graph, let $r_0$ be an arbitrary vertex of $G$, let $G^+$ be a graph obtained by adding a dummy vertex at each crossing in $G$ and then arbitrarily adding non-crossing edges to obtain a triangulation.  Then $G$ has a layered $H$-partition $(\mathcal{L}=\langle V_0',V_1',\ldots\rangle,\mathcal{P})$ satisfying the conditions of \cref{k-planar} and such that $V_i' := \{v\in V(G): \floor{\dist_{G^+}(r_0,v)/(k+1)}=i\}$ for each integer $i\ge 0$.

@@

%  which leads to the layering $\langle V_0,V_1,\ldots\rangle$ of $G^+$ where $V_i=\{v\in V(G^+): \dist_{T}(r,v) = i\}$.

@@

%  The rest of proof remains unchanged until the point where we derive the layering $\mathcal{L}=\langle V_0',V_1',\ldots\rangle$ of $G$.  At this point, we set

@@

%    V_i' := \{v\in V(G): \floor{\dist_{T}(r,v)-2}=i\} 

@@

%   Let $G$ be an $(\Sigma,k)$-plane graph where $\Sigma$ is a surface of Euler genus $g$.  Let $G^+$ be the $(\Sigma,0)$-plane graph obtained by adding a dummy vertex at each crossing in $G$.  Let $r_0$ be any vertex in $V(G)$, let $T$ be BFS spanning tree of $G^+$ rooted at $r_0$ and let $\langle V_0,V_1,\ldots\rangle$ be the resulting layering of $G^+$.

@@

%   Since $|Z\cap V_i| \le 2g$, the same argument used in the proof of \cref{width-of-g-plus} shows that $|Z'\cap V_i|\le 2g\times 2\times (2k+1) = 4g(2k+1)$ for each integer $i\ge 0$.

@@

%   Apply \cref{cor-k-planar} or (in the case $k=1$) \cref{cor-1-planar} to $G^*_1$ using $G^{*+}_1$ and the distinguished vertex $r^*$ to obtain a layered $H$-partition of $G^*_1$.  Since $G^*_1\supseteq G^*_0$ and $G^*_0$ is isomorphic to $G-Z'$, this gives a layered $H$-partition $(\mathcal{L}=\langle V_0',V_1',\ldots\rangle,\mathcal{P})$ of $G-Z'$ that satisfies the conditions of \cref{cor-k-planar} or \cref{cor-1-planar}, as appropriate.  In particular, for every integer $i\ge 0$,

@@

%     V_i'& = \{v\in V(G-Z'):\floor{\dist_{G^*}(r^*,v)/(k+1)} = i\} \\

@@

%   It follows that $|V_i'\cap Z'|\le 4g(2k+1)(k+1) = 8gk^2+12gk+4g$.  

@@

%     |V_i' \cap P| \le \begin{cases} 8gk^2+12gk+4g & \text{if $P=Z'$} \\

>>>>

\documentclass{patmorin}
\listfiles
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}

@@

% \usepackage[mathlines]{lineno}
% \setlength{\linenumbersep}{2em}
% \linenumbers
% \rightlinenumbers
% \linenumbers
% \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
%  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
%  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
%  \renewenvironment{#1}%
%     {\linenomath\csname old#1\endcsname}%
%     {\csname oldend#1\endcsname\endlinenomath}}%
% \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
%  \patchAmsMathEnvironmentForLineno{#1}%
%  \patchAmsMathEnvironmentForLineno{#1*}}%
% \AtBeginDocument{%
% \patchBothAmsMathEnvironmentsForLineno{equation}%
% \patchBothAmsMathEnvironmentsForLineno{align}%
% \patchBothAmsMathEnvironmentsForLineno{flalign}%
% \patchBothAmsMathEnvironmentsForLineno{alignat}%
% \patchBothAmsMathEnvironmentsForLineno{gather}%
% \patchBothAmsMathEnvironmentsForLineno{multline}%
% }
%
%
% \allowdisplaybreaks
%\sloppy
%%%%%%%%%%
% \makeatletter
% \def\NAT@spacechar{~}
% \makeatother

\newcommand{\snote}[1]{{\color{red}#1}}

\crefname{lem}{Lemma}{Lemmas}
\crefname{thm}{Theorem}{Theorems}
\crefname{cor}{Corollary}{Corollaries}
\crefname{conj}{Conjecture}{Conjectures}
\crefname{obs}{Observation}{Observations}
\crefname{prop}{Proposition}{Propositions}
\crefname{clm}{Claim}{Claims}
\crefname{openproblem}{Open Problem}{Open Problems}
\crefformat{equation}{(#2#1#3)}
\Crefformat{equation}{Equation #2(#1)#3}

@@

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and implicit graph encoding. We extend this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest.
\end{abstract}
\end{titlepage}
\pagenumbering{roman}
\tableofcontents
\newpage

\pagenumbering{arabic}
% \setcounter{\page}{1}
\section{Introduction}
\label{Introduction}

@@

A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle L_0,L_1,\ldots\rangle$ such that $\{L_0,L_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in L_i$ and $w\in L_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.

@@

Apex-minor-free graphs, addressed by \cref{ApexMinorFree}, are the largest minor-closed class that admit bounded layered partitions~\citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. \snote{DCG Reviewer \#1 writes: \emph{Isn't this trivial? A grid with all
diagonals will do.}  Are they correct? If so, what to do?} Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

@@

\begin{lem}\label{generalized-tripod}
  Let $G$ be a graph; let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$ where $H$ has treewidth at most $t$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \lemref{nice-decomposition}.  For each $x\in V(T)$, let $V_x := \bigcup_{y\in V(T_x)} Y_y$, $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x:=V_x\cup F_x$.  Then,
  \begin{compactenum}[(Y1)]
    % \item\ylabel{thickness} $\mathcal{Y}=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width at most $\ell$ with respect to $\mathcal{L}$.
    \item\ylabel{separator} For each $x\in V(T)$, there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$.
    \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{lem}

@@

Next we show that $\mathcal{S}$ has small layered width with respect to the layering $\mathcal{L}=\langle L_1,\ldots,L_h\rangle$:

\begin{clm}\clmlabel{general-width}
  For each $i\in\{1,\ldots,h\}$ and each $x\in V(J)$, $|S_x\cap L_i|\le d\ell(k^2+3)$.
\end{clm}

\begin{proof}
  We count the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each vertex $w\in Y_x$. Refer to \cref{contribute}.  If $w\in Y_x\cap L_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap L_i$.

  \begin{figure}[htbp]
    \begin{center}
      \includegraphics{figs/contribute}
    \end{center}
    \caption{A path $P$ containing an internal vertex $w\in Y_x\cap L_{i-j}$.}
    \label{contribute}
  \end{figure}

  Otherwise, consider some path $P\in\mathcal{P}$ that contains $w$ as an internal vertex.  If $w\in L_{i}$, then $P$ contributes at most $k+1$ vertices to $S_x\cap L_i$.  If $w\in L_{i-1}\cup L_{i+1}$, then $P$ contributes at most $k$ vertices to $S_x\cap L_i$. If $w\in L_{i-j}\cup L_{i+j}$ for $j\ge 2$, then $P$ contributes at most $k-j$ vertices to $S_x\cap L_i$.

  For any $j$, the number of vertices $w\in L_{i+j}\cap Y_x$ is at most $\ell$. Each such vertex $w$ is an internal vertex of at most $d$ paths in $\mathcal{P}$. Therefore,
  \[  |S_x\cap L_i|\le d\ell\cdot\left(k+1 + 2k + \sum_{j=2}^k 2(k-j)\right) %= d(2k+1) + \sum_{i=1}^{k-2} i
      = d\ell(k^2 +3) \enspace . \qedhere
  \]
\end{proof}

@@

  At this point, the proof of \cref{modern-major-general} is almost immediate from \clmref{general-width} and \clmref{general-bag-size}, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain edges $vw$ with $v\in L_i$ and $w\in L_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle L_0',\ldots,L_h'\rangle$ in which $L_i'=\bigcup_{j=ki}^{ki+k-1} L_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+3)$ to $d\ell(k^3+3k)$.  Therefore $G$ has an $H$-partition of layered width at most $d\ell(k^3+3k)$ in which $H$ has treewidth at most $\binom{k+t}{t}$, completing the proof of \cref{modern-major-general}.
\end{proof}

@@

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $L_i\cap Y_x$.  More precisely, each path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (L_{i-j}\cup L_{i+j})$ contributes: (i)~at most two vertices to $S_x\cap L_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}\}$; (ii)~at most one vertex to $S_x\cap L_j$ for $j\in\{\floor{(k+1)/2}+1,\ldots,k+1\}$; or (iii)~no vertices to $S_x\cap L_j$ for $j > k+1$.
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields
\begin{align*}
    |S_x\cap Y_i| \le d\ell\left(
       2
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor
       + 2\left\lceil\frac{k+1}{2}\right\rceil
       \right)
       & =
       d\ell\left(
         2 + 2(k+1) + 2\left\lfloor\frac{k+1}{2}\right\rfloor
       \right) \\
       & \le
       d\ell(3k+5)
      = 18k+30 \enspace .
\end{align*}
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $(18k+30)(k+1)=18k^2+48k+30$.  This establishes \cref{k-planar}.
% Note to self-and-others. In our original SODA submission we had 18k^2+30k here because we forgot that the layer compression step has compress k+1 consecutive layers, not just k consecutive layers.

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analagous result for plane graphs \cite{dujmovic.joret.ea:planar}.
\begin{lem}
\label{induction} The setup:
  \begin{compactenum}
    \item Let $G$ and $G'$ be defined as above.
    \item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
    \item For each integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
    \item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
    \begin{compactenum}
      \item No edge of $F$ is crossed by any edge of $G$; and
      \item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $F[P_i]$ is a path; and
        \item $|V(P_i)\cap L_j| \le 15$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
  \end{compactenum}
  Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
  \begin{compactenum}
    \item $H$ is planar;
    \item for all integers $j\ge 0$, and all $x\in V(H)$, $|S_x\cap L_j|\le 15$;
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$; and
    \item $H$ has a tree decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

@@

  We claim that, for each integer $i\ge 0$, $|V(\overline{Y}^+)\cap L_i|\le 15$.  First observe that, since $Q_1,Q_2,Q_3$ are each vertical paths in $T$,  $\overline{Y}$ contains at most three vertices of $L_i$, each incident on at most two edges of $\overline{Y}$.  Since $\dist_{G'}(v,w)\le 2$ for each $vw\in E(G)$, any vertex $x\in V(\overline{Y}^+)\setminus V(\overline{Y})\cap L_i$, is incident to an edge $xy\in E(G)$ that crosses one of the at most six edges in $\overline{Y}$ having an endpoint in $L_i$.  These at most six edges have at most 12 endpoints.  Therefore $|V(\overline{Y}^+)\setminus V(\overline{Y})\cap L_i|\le 6\times 2=12$, so $|V(\overline{Y}^+)\cap L_i|\le 12+3=15$.

@@

  Use the layering $\mathcal{L}=\langle L_0',L_1'\ldots\rangle$ where $L_i'=L_{2i}\cup L_{2i+1}$ for each integer $i\ge 0$. That this is a layering of $G$ follows from the fact that $\dist_{G'}(v,w)\le 2$ for every edge $vw\in E(G)$.  Since $|L_j\cap S_x|\le 15$ for every integer $j\ge 0$, $|L_i'\cap S_x|\le 30$ for every integer $i\ge 0$ and every $x\in V(H)$.
\end{proof}

% Exactly the same argument used to establish \cref{cor-k-planar} as a consequence of \cref{k-planar} establishes the following result as a consequence of \cref{1-planar}:
%
% \begin{cor}
% \label{cor-1-planar}
% Let $G$ be a $1$-plane graph, let $r_0$ be an arbitrary vertex of $G$, let $G^+$ be a graph obtained by adding a dummy vertex at each crossing in $G$ and then arbitrarily adding non-crossing edges to obtain a triangulation.  Then $G$ has a layered $H$-partition $(\mathcal{L}=\langle L_0',L_1',\ldots\rangle,\mathcal{P})$ satisfying the conditions of \cref{1-planar} and such that $L_i' = \{v\in V(G): \floor{\dist_{G^+}(r_0,v)/2}=i\}$ for each integer $i\ge 0$.
% \end{cor}

@@

%Let $G$ be a $k$-plane graph, let $r_0$ be an arbitrary vertex of $G$, let $G^+$ be a graph obtained by adding a dummy vertex at each crossing in $G$ and then arbitrarily adding non-crossing edges to obtain a triangulation.  Then $G$ has a layered $H$-partition $(\mathcal{L}=\langle L_0',L_1',\ldots\rangle,\mathcal{P})$ satisfying the conditions of \cref{k-planar} and such that $L_i' := \{v\in V(G): \floor{\dist_{G^+}(r_0,v)/(k+1)}=i\}$ for each integer $i\ge 0$.

@@

%  which leads to the layering $\langle L_0,L_1,\ldots\rangle$ of $G^+$ where $L_i=\{v\in V(G^+): \dist_{T}(r,v) = i\}$.

@@

%  The rest of proof remains unchanged until the point where we derive the layering $\mathcal{L}=\langle L_0',L_1',\ldots\rangle$ of $G$.  At this point, we set

@@

%    L_i' := \{v\in V(G): \floor{\dist_{T}(r,v)-2}=i\}

@@

%   Let $G$ be an $(\Sigma,k)$-plane graph where $\Sigma$ is a surface of Euler genus $g$.  Let $G^+$ be the $(\Sigma,0)$-plane graph obtained by adding a dummy vertex at each crossing in $G$.  Let $r_0$ be any vertex in $V(G)$, let $T$ be BFS spanning tree of $G^+$ rooted at $r_0$ and let $\langle L_0,L_1,\ldots\rangle$ be the resulting layering of $G^+$.

@@

%   Since $|Z\cap L_i| \le 2g$, the same argument used in the proof of \cref{width-of-g-plus} shows that $|Z'\cap L_i|\le 2g\times 2\times (2k+1) = 4g(2k+1)$ for each integer $i\ge 0$.

@@

%   Apply \cref{cor-k-planar} or (in the case $k=1$) \cref{cor-1-planar} to $G^*_1$ using $G^{*+}_1$ and the distinguished vertex $r^*$ to obtain a layered $H$-partition of $G^*_1$.  Since $G^*_1\supseteq G^*_0$ and $G^*_0$ is isomorphic to $G-Z'$, this gives a layered $H$-partition $(\mathcal{L}=\langle L_0',L_1',\ldots\rangle,\mathcal{P})$ of $G-Z'$ that satisfies the conditions of \cref{cor-k-planar} or \cref{cor-1-planar}, as appropriate.  In particular, for every integer $i\ge 0$,

@@

%     L_i'& = \{v\in V(G-Z'):\floor{\dist_{G^*}(r^*,v)/(k+1)} = i\} \\

@@

%   It follows that $|L_i'\cap Z'|\le 4g(2k+1)(k+1) = 8gk^2+12gk+4g$.

@@

%     |L_i' \cap P| \le \begin{cases} 8gk^2+12gk+4g & \text{if $P=Z'$} \\
